Generalised gravitational burst generation with Generative Adversarial Networks

### 2020-09-14 Paper_draft/main.tex
edits to plots

\begin{table}[hb]
\centering
\caption{Burst training parameters~\chris{Add a bit more caption description
here. Plus address the issues of the actual distributions used (uniform) and
how you include the BBH parameters. It's also not obvious that all the classes
share the same 3 parameters.}}
%\footnotesize
\begin{tabular}{@{} l l l l l l }
\br
\hline
 Waveform & Central frequency  & Decay & Central time epoch & Mass range \\
 & (Hz) & (s) & (s) & ($\textrm{M}_{\odot}$) \\
\mr
Sine-Gaussian & 70 - 250 & 0.004 - 0.03 & 0.4 - 0.6 & N/A  \\  
Ringdown & 70 - 250 & 0.004 - 0.03 & 0.4 - 0.6 & N/A \\
White-noise burst & 70 - 250 & 0.004 - 0.03 & 0.4 - 0.6 & N/A  \\
Gaussian pulse & N/A & 0.004 - 0.03 & 0.4 - 0.6 & N/A  \\
BBH & N/A & N/A & N/A & 5 - 70  \\
 \br
\end{tabular}\\
\label{Tab:training_parms}
\end{table}
\normalsize

@@

% list the 5 waveform classes
%
\begin{itemize}
%
\item {\bf Sine-Gaussian}: $h_{\text{sg}}(t) = A \exp\left[ - (t-t_{0})^2 /
\tau^2 \right] \sin (2 \pi f_0 (t-t_0))$, a sinusoidal wave with a Gaussian
envelope characterised by a central frequency, $f_0$, amplitude, $A$, time of arrival, $t_{0}$. 
%
\item {\bf Ring-down}: $h_{\text{rd}}(t) = A \exp \left[-{(t-t_0)} / {\tau}
\right] \sin(2 \pi f_0 (t-t_0))$, with frequency $f_0$ and duration $\tau$, amplitude, $A$, time of arrival, $t_{0}$. 
%
\item {\bf White-noise bursts}: $h_{\text{wn}}(t_j) = Ag_j\exp\left[ -
(t-t_{0})^2 / \tau^2 \right]$ where $g_j$ are drawn from a zero mean unit
variance Gaussian distribution with a Gaussian envelope of duration $tau$.
%
\item {\bf Gaussian pulse}: $h_{\text{gp}}(t) = \exp(-t^2 / \tau^2)$ with
duration parameter $\tau$.
%
\item {\bf Binary-black hole}: Simulated using the IMRPhenomD
waveform~\cite{Khan_2016} routine from LALSuite~\cite{lalsuite} which models
the inspiral, merger and ringdown of a \ac{BBH} waveform. The component masses
lie in the range of [5,70] $\textrm{M}_{\odot}$ with zero spins and we fix
$m_1>m_2$. The mass distribution is approximated by a power law with
index of 1.6~\cite{Abbott_2019}. The inclinations are drawn
such that the cosine of the angles lies uniformly in the range [-1,1] and we only use the plus polarisation.
%
\end{itemize}
%
The location of the peak amplitude of the waveforms (corresponding to the
mid-points of all but the ring-down and \ac{BBH} classes) are randomly drawn from a uniform distribution to
be within [0.4, 0.6] sec from the start of the 1 sec time interval and all
training waveforms are sampled at 1024 Hz.  The parameter prior ranges are
defined in~\cref{Tab:training_parms}. All training data is rescaled such that their amplitudes peak at 1.

@@

% describe the main idea
%
Given a 100-dimensional vector drawn from a normal distribution and a class
label, the GAN is able to generate burst-like waveforms generalised from the
training set~\chris{please state this in terms of the latent space and class
space}. We set out by describing the quality of generated waveforms and
how they compare to the training set. We then explore the structure of the
latent and class spaces by interpolating between points in these spaces. We
test two methods of sampling from the class space that can be used to generate
a new breed of signal composed of weighted elements of each training class.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/generations/sg.png}
    \includegraphics[width=\textwidth]{figures/generations/rd.png}
    \includegraphics[width=\textwidth]{figures/generations/wnb.png}
    \includegraphics[width=\textwidth]{figures/generations/blip.png}
    \includegraphics[width=\textwidth]{figures/generations/bbh.png}
    \caption{Examples of \ac{GW} burst signals generated by a conditional generative adversarial network. By row: Sine-Gaussian, Ringdown,
White-noise burst, Gaussian pulse, Binary black hole merger.~\chris{please add
lots more description to the caption regarding wjhat you are plotting. Mention
that these are generated waveforms plotted as a function of time, latent space
values are randomised for each panel (or are they the same for each column?)
classes on each row. Also mention the zoom in on the time axis. Also fix the
time axis limits so they are consistent across all waveforms.}}
\label{fig:gen_signals} 
\end{figure}

@@

The generator network is a function $G$ : $\mathbf{z},\mathbf{c} $ $\in$
$\mathbb{R}^{100}$ $\to$ $\mathbb{R}^{1024\times2}$,~\chris{why is this
outputting to 2 timeseries? Is this out of date?} where $\mathbf{z},\mathbf{c}$
are the latent vector and class vector~\chris{these have already been defined
}. Given a latent vector randomly sampled from a normal distribution with zero
mean and unit variance, a class label which is represented by a 5 dimensional
one-hot encoded vector for each class~\chris{give an example e.g.
$c=[0,1,0,0,0]$ represents the ring-down signal class}, the results from the
generator can be seen in \cref{fig:gen_signals}~\chris{please indicate that
this is all done after training has completed}. Each plot shows the output of
the generator for random input latent space vectors $\mathbf{z}$ and each row
corresponds to one of the five class vectors $\mathbf{c}$.~\chris{yes? What
about the plot? Anything positive to say about what you're showing? Please
comment on the quality of the generated waveforms and the good varaition within
each class representative of the variation in the training data. The point is
to show the general variation and quality.}

@@

% expain what we are planning to show here
%
\ac{ML} algorithms are often described as universal function approximators. In
the case of the generator it maps samples drawn from a 100 dimensional Gaussian
distribution in the latent space to its representation of the training set. As
with any function, there should be a one to one mapping from the domain and
co-domain to allow for smooth transitions across the latent space. One
advantage of using \acp{GAN} as a waveform generator is that once it is
trained, it can perform rapid generations faster than computationally expensive
algorithms \michael{e.g.?}~\chris{you can maybe reference Seb Kahn and Green
here soon. It will be on the arXiv next week I think - they did fast CBC
waveform generation.  Also Walter Del Pozzo has a paper xoming out soon on
this. Neither use GANs though, just ML}. For complicated data sets, the network
architecture must be diverse and dense enough to capture distinct variations
from the training set.  Most GANs perform well on relatively low resolution
image generations, however, higher resolutions demand larger networks and long
training times. GANs attempting to replicate complicated structures and do not
have the necessary architecture either struggle to produce results at all or
fall into the common failure mode know as mode collapse; where the generator
produces a small variety of samples or simply memorises the training set. To
test this, we perform linear interpolations in the latent and class
space.~\chris{lots of technical talk before actually saying what you are trying
to do. Reverse this. Also try to provide references for the technical talk}

\chris{this paragraph should come before the other one since it explains what
you are doing before going into technical stuff} In this section we explore the
latent space formed by the generator by interpolation. We take two random
points in the latent space and linearly interpolate between them~\chris{OK, but
how do you select the latent points? Are they the same pair of points for each
class?}. These new latent space vectors can now be fed into the generator to
make predictions on~\chris{"to make predictions on" - what does that mean?}
while keeping the class vectors constant at each of the 5 \chris{fiducial?}
training classes. The full effect shown in~\cref{fig:z_interp}~\chris{not a
sentence}. We can see that each plot shows plausible waveforms suggesting that
the generator has constructed a smooth space unlike the discrete training
case~\chris{don't confuse the issue by saying that the class space is discrtee
when later we want to show the opposite}. Additionally as each class is given
the same latent points to interpolate over, we can see that the waveforms
cluster together with respect to their parameters. Visually, the sine-gaussian
and ring-down waveforms share similar frequencies and the other signals show
similar decays and starting epochs for a common latent space input. The only
exception is for the \ac{BBH} waveforms. This can be explained by the fact that
they were trained with more variety of parameters~\chris{what is more variety
of parameters? Does it really mean anything?} and consistently have their peak
amplitudes in the last quarter of the time series~\chris{do they? In the plot
the peak amplitude locations seem to range from 0.6 back to 0.5 but that's not
too different to the ring-downs, the white noise or the pulse. The
sine-Gaussian is the one that stands out as different. Can you please verify
the correctness of the plot you are referring to.}.

@@

>>>>

\begin{table}[hb]
\centering
\caption{The parameters used in generating the set of training data. The parameters are drawn uniformly in the below ranges. ~\chris{Add a bit more caption description
here. Plus address the issues of the actual distributions used (uniform) and
how you include the BBH parameters. It's also not obvious that all the classes
share the same 3 parameters.}}
%\footnotesize
\begin{tabular}{@{} l l l l l l }
\br
\hline
 Waveform & Central frequency  & Decay & Central time epoch & Mass range \\
 & (Hz) & (s) & (s) & ($\textrm{M}_{\odot}$) \\
\mr
Sine-Gaussian & 70 - 250 & 0.004 - 0.03 & 0.4 - 0.6 & N/A  \\  
Ringdown & 70 - 250 & 0.004 - 0.03 & 0.4 - 0.6 & N/A \\
White-noise burst & 70 - 250 & 0.004 - 0.03 & 0.4 - 0.6 & N/A  \\
Gaussian pulse & N/A & 0.004 - 0.03 & 0.4 - 0.6 & N/A  \\
BBH & N/A & N/A & N/A & 5 - 70  \\
 \br
\end{tabular}\\
\label{Tab:training_parms}
\end{table}
\normalsize

@@

% list the 5 waveform classes
%
\begin{itemize}
%
\item {\bf Sine-Gaussian}: $h_{\text{sg}}(t) = A \exp\left[ - (t-t_{0})^2 /
\tau^2 \right] \sin (2 \pi f_0 (t-t_0))$, a sinusoidal wave with a Gaussian
envelope characterised by a central frequency, $f_0$, amplitude, $A$, time of arrival, $t_{0}$. 
%
\item {\bf Ring-down}: $h_{\text{rd}}(t) = A \exp \left[-{(t-t_0)} / {\tau}
\right] \sin(2 \pi f_0 (t-t_0))$, with frequency $f_0$ and duration $\tau$, amplitude, $A$, time of arrival, $t_{0}$. 
%
\item {\bf White-noise bursts}: $h_{\text{wn}}(t_j) = Ag_j\exp\left[ -
(t-t_{0})^2 / \tau^2 \right]$ where $g_j$ are drawn from a zero mean unit
variance Gaussian distribution with a Gaussian envelope of duration $tau$.
%
\item {\bf Gaussian pulse}: $h_{\text{gp}}(t) = \exp(-t^2 / \tau^2)$ with
duration parameter $\tau$.
%
\item {\bf Binary-black hole}: Simulated using the IMRPhenomD
waveform~\cite{Khan_2016} routine from LALSuite~\cite{lalsuite} which models
the inspiral, merger and ringdown of a \ac{BBH} waveform. The component masses
lie in the range of [5,70] $\textrm{M}_{\odot}$ with zero spins and we fix
$m_1>m_2$. The mass distribution is approximated by a power law with
index of 1.6~\cite{Abbott_2019}. The inclinations are drawn
such that the cosine of the angles lies uniformly in the range [-1,1] and we only use the plus polarisation.
%
\end{itemize}
%
The location of the peak amplitude of the waveforms (corresponding to the
mid-points of all but the ring-down and \ac{BBH} classes) are randomly drawn from a uniform distribution to
be within [0.4, 0.6] sec from the start of the 1 sec time interval and all
training waveforms are sampled at 1024 Hz.  The parameter prior ranges are
defined in~\cref{Tab:training_parms} and a sample of training waveforms are shown in \ref{fig:training_waveforms}. All training data is rescaled such that their amplitudes peak at 1.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{Paper_draft/figures/training-4.png}
    \caption{Examples of the five different waveforms that were used in training the \ac{GAN} for this study. \jordan{put in the parms here too?}}
    \label{fig:training_waveforms}
\end{figure}

@@

% describe the main idea
%
Given a 100-dimensional vector drawn from a normal distributed latent space and a one-hot encoded class
label, the GAN is able to generate burst-like waveforms generalised from the
training set. We set out by describing the quality of generated waveforms and
how they compare to the training set. We then explore the structure of the
latent and class spaces by interpolating between points in these spaces. We
test two methods of sampling from the class space that can be used to generate
a new breed of signal composed of weighted elements of each training class.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/generations/sg.png}
    \includegraphics[width=\textwidth]{figures/generations/rd.png}
    \includegraphics[width=\textwidth]{figures/generations/wnb.png}
    \includegraphics[width=\textwidth]{figures/generations/blip.png}
    \includegraphics[width=\textwidth]{figures/generations/bbh.png}
    \caption{\ac{GAN} Generated waveforms plotted as a function of time. The latent space inputs for each panel is randomised and each row is assigned one of the five class vectors. By row: Sine-Gaussian, Ringdown,
White-noise burst, Gaussian pulse, Binary black hole merger. On the first four rows each x axis is zoomed for ease of viewing, while the last row shows the full one second. \jordan{re-do with latent space all zeros to all ones (see colab)}}
\label{fig:gen_signals} 
\end{figure}

@@

After training the networks, the generator's function is to map $G$ : $\mathbf{z},\mathbf{c} $ $\in$
$\mathbb{R}^{100}$ $\to$ $\mathbb{R}^{1024}$. The results from the
generator can be seen in \cref{fig:gen_signals}. Each plot shows the output of
the generator for random input latent space vectors $\mathbf{z}$ and each row
corresponds to one of the five class vectors $\mathbf{c}$.~\chris{yes? What
about the plot? Anything positive to say about what you're showing? Please
comment on the quality of the generated waveforms and the good varaition within
each class representative of the variation in the training data. The point is
to show the general variation and quality.}

@@

% expain what we are planning to show here
%
In this section we explore the latent space formed by the generator by interpolation. We take two random
points in the latent space and linearly interpolate between them~\chris{OK, but
how do you select the latent points? Are they the same pair of points for each
class?}. These new latent space vectors can now be fed into the generator to
make predictions on~\chris{"to make predictions on" - what does that mean?}
while keeping the class vectors constant at each of the 5 \chris{fiducial?}
training classes. The full effect shown in~\cref{fig:z_interp}~\chris{not a
sentence}. We can see that each plot shows plausible waveforms suggesting that
the generator has constructed a smooth space unlike the discrete training
case~\chris{don't confuse the issue by saying that the class space is discrtee
when later we want to show the opposite}. Additionally as each class is given
the same latent points to interpolate over, we can see that the waveforms
cluster together with respect to their parameters. Visually, the sine-gaussian
and ring-down waveforms share similar frequencies and the other signals show
similar decays and starting epochs for a common latent space input. The only
exception is for the \ac{BBH} waveforms. This can be explained by the fact that
they were trained with more variety of parameters~\chris{what is more variety
of parameters? Does it really mean anything?} and consistently have their peak
amplitudes in the last quarter of the time series~\chris{do they? In the plot
the peak amplitude locations seem to range from 0.6 back to 0.5 but that's not
too different to the ring-downs, the white noise or the pulse. The
sine-Gaussian is the one that stands out as different. Can you please verify
the correctness of the plot you are referring to.}.

@@

\ac{ML} algorithms are often described as universal function approximators. In
the case of the generator it maps samples drawn from a 100 dimensional Gaussian
distribution in the latent space to its representation of the training set. As
with any function, there should be a one to one mapping from the domain and
co-domain to allow for smooth transitions across the latent space. One
advantage of using \acp{GAN} as a waveform generator is that once it is
trained, it can perform rapid generations faster than computationally expensive
algorithms \michael{e.g.?}~\chris{you can maybe reference Seb Kahn and Green
here soon. It will be on the arXiv next week I think - they did fast CBC
waveform generation.  Also Walter Del Pozzo has a paper xoming out soon on
this. Neither use GANs though, just ML}. For complicated data sets, the network
architecture must be diverse and dense enough to capture distinct variations
from the training set.  Most GANs perform well on relatively low resolution
image generations, however, higher resolutions demand larger networks and long
training times. GANs attempting to replicate complicated structures and do not
have the necessary architecture either struggle to produce results at all or
fall into the common failure mode know as mode collapse; where the generator
produces a small variety of samples or simply memorises the training set. To
test this, we perform linear interpolations in the latent and class
space.~\chris{lots of technical talk before actually saying what you are trying
to do. Reverse this. Also try to provide references for the technical talk}
